Ecological niche models of plant species exploited by Epipalaeolithic foragers

### 2023-04-26 analysis/analysis.Rmd
Updates from Feb 2020

---
title: Ecological niche models of plant species exploited by Epipalaeolithic foragers
  in the *Badia* (c. 14500 BP)
output:
  html_document:
    df_print: paged
  pdf_document:
    latex_engine: xelatex
---

## Dependencies

```{r packages, warning=FALSE, message=FALSE}
library("tidyverse")
library("magrittr")
library("sf")
library("raster")
library("broom")
library("dismo")
library("rnaturalearth")
library("gt")
library("BadiaPaleoFloraENM")
```

@@

```{r regions, warning=FALSE, message=FALSE}
sw_asia <- extent(c(30, 50, 25, 40))
s_levant <- extent(c(34, 39, 29, 34))
palestine <- ne_countries(scale = 10, country = c("Israel", "Palestine"),
             returnclass = "sf") %>% st_union() %>% st_sf()

land <- read_sf("data/raw_data/ne/ne_10m_land.shp")[,"featurecla"] %>%
  st_crop(sw_asia)
rivers <- read_sf("data/raw_data/ne/ne_10m_rivers_lake_centerlines_scale_rank.shp") %>% 
  st_crop(s_levant)
water <- rbind(read_sf("data/raw_data/ne/ne_10m_ocean.shp")[,"featurecla"],
               read_sf("data/raw_data/ne/ne_10m_lakes.shp")[,"featurecla"]) %>% 
  st_crop(sw_asia)
```

```{r regions-map, echo=FALSE, results='hide'}
basemap <- ggplot() + 
  geom_sf(data = land, fill ="grey") +
  geom_sf(data = palestine, fill = "lightgrey") + 
  geom_sf(data = rivers, colour = "white") +
  geom_sf(data = water, fill = "white") +
  coord_sf(xlim = c(xmin(s_levant), xmax(s_levant)),
           ylim = c(ymin(s_levant), ymax(s_levant)))
basemap
```

## Occurrence data

@@

```{r occurrence-data, warning=FALSE, message=FALSE}
# Download or read data from GBIF
gbif_path <- here::here("analysis/data/raw_data/gbif")
occ <- bind_rows(
  fetch_gbif("Hordeum", "spontaneum*", ext = sw_asia, gbif_path),
  fetch_gbif("Bolboschoenus", "glaucus*", ext = sw_asia, gbif_path),    
  fetch_gbif("Bolboschoenus", "maritimus*", ext = sw_asia, gbif_path),    
  fetch_gbif("Scirpus", "maritimus*", ext = sw_asia, gbif_path),
  fetch_gbif("Triticum", "urartu*", ext = sw_asia, gbif_path),
  fetch_gbif("Triticum", "dicoccoides*", ext = sw_asia, gbif_path),
  fetch_gbif("Stipa", "*", ext = sw_asia, gbif_path),
  fetch_gbif("Zilla", "spinosa*", ext = sw_asia, gbif_path),
  fetch_gbif("Vitex", "*", ext = sw_asia, gbif_path),
  fetch_gbif("Fraxinus", "*", ext = sw_asia, gbif_path)
)

# Code taxonomic groups for analysis
occ %>% 
  mutate(taxon = case_when(genus == "Hordeum" ~ "Hordeum spontaneum",
                           genus == "Bolboschoenus" ~ "Bolboschoenus sp. (=Scirpus maritimus)",
                           species == "Triticum urartu" ~ "Triticum urartu",
                           species == "Triticum dicoccoides" ~ "Triticum dicoccoides",
                           genus == "Stipa" ~ "Stipa sp.",
                           genus == "Zilla" ~ "Zilla spinosa",
                           genus == "Vitex" ~ "Vitex sp.",
                           genus == "Fraxinus" ~ "Fraxinus sp.")) %T>%
  {if(any(is.na(.$taxon))) stop("Taxon contains missing values!")} ->
  occ

@@

It's interesting to map the current distribution of these plants.
```{r occurrence-map, echo=FALSE, results='hide'}
ggplot(occ, aes(x = lon, y = lat)) +
  geom_sf(data = occ, shape = 1, size=0.5) +
  geom_sf(data = water, inherit.aes = FALSE, fill = "white", colour = "darkgrey") +
  facet_wrap(~taxon)
```

## Pseudo-absences

Occurrence data only tells us where a species is present;
there is no definitive information on where the species is *not* found.
We therefore need to generate random background points or "pseudo-absences" to feed to the model.
There are several ways to do this. 
We follow the advice of Barbet-Massin et al. (2012) for regression-based species distribution models and simply use a large (~10000) random sample of points.

```{r psabsences, warning=FALSE, message=FALSE}
occ <- as_tibble(occ)
occ$occurrence <- TRUE
for(taxon in unique(occ$taxon)) {
  occ <- bind_rows(occ, generate_psabsences(palestine, taxon = taxon,
                                            occurrence = FALSE, N = 10000))
}
occ <- sf::st_as_sf(occ, crs = 4326)
rm(taxon)
```

## Environmental data

### Bioclimatic variables

@@

```{r bioclim-table, echo=FALSE}
tibble(biovar = paste0("bio", 1:16),
       definition = c("Mean annual temperature",
                      "Mean diurnal range",
                      "Isothermality",
                      "Temperature seasonality",
                      "Maximum temperature of warmest month",
                      "Minimum temperature of coldest month",
                      "Annual temperature range",
                      "Mean temperature of wettest quarter",
                      "Mean temperature of driest quarter",
                      "Mean temperature of warmest quarter",
                      "Mean temperature of coldest quarter",
                      "Total annual precipitation",
                      "Precipitation of wettest month",
                      "Precipitation of driest month",
                      "Precipitation seasonality",
                      "Precipitation of wettest quarter"),
       formula = c("",
                   "",
                   "(bio2 / bio7) * 100",
                   "sd(temp) * 100",
                   "",
                   "",
                   "bio5 - bio6",
                   "",
                   "",
                   "",
                   "",
                   "",
                   "",
                   "",
                   "",
                   "")
) %>%
  gt()
```

Here the climate data is from WorldClim 1.4 (<http://www.worldclim.org>), a global dataset compiled from weather station records from the period 1960–1990.
WorldClim has been widely used in ecological niche modelling.
Other global climate datasets (e.g. CHELSA, MERRAClim) and a newer version of WorldClim are now available, so it may be a good idea to investigate using these instead.
The WorldClim data is large so again it is cached in the package's `derived_data` directory.

```{r bioclim-data, echo=FALSE}
# bioclim <- get_bioclim(s_levant, resolution = 0.5, cache_path = "data/derived_data")
# Workaround for memory problems in Safawi
load("data/derived_data/bioclim.RData")
plot(bioclim)
```

### Soil variables

@@

```{r soil-data}
soilgrids <- raster::stack(list.files("data/raw_data/soilgrids", ".*tif", full.names = TRUE))
plot(soilgrids)
```

@@

There are also relevent aspects of the terrain that can be easily derived from a digital elevation model.
We'll use NASA-JPL's SRTM v3 1 arc-second resolution DEM.

### Selection of predictor variables

@@

```{r review-table, echo=FALSE, warning=FALSE, message=FALSE}
read_csv("data/raw_data/flora_review.csv") %>% 
  group_by(Taxon) %>% 
  gt()
```

@@

```{r predictor-table, echo=FALSE}
pred <- bind_rows(
  tibble(group = rep("Climate", 5),
         variable = c("Minimum temperature of coldest month",
                      "Total annual precipitation", "Precipitation of wettest month",
                      "Precipitation of driest month", "Precipitation seasonality"),
         units = c("°C ×10", "mm", "mm", "mm", "σ/μ"),
         source = rep("WorldClim 1.4", 5),
         code = c("bio6", "bio12", "bio13", "bio14", "bio15"),
         raster = as.list(raster::subset(bioclim,
                                         c("bio6", "bio12", "bio13", "bio14", "bio15")))),
  tibble(group = rep("Soil", 5),
         variable = c("Organic carbon content", "Coarse fragment content",
                      "Sand content", "Acidity", "Depth to bedrock"),
         units = c("g per kg fine earth", "% volume", "% mass", "pH ×10 in H2O",
                   "cm (up to 200 cm)"),
         source = rep("SoilGrids250m", 5),
         code = c("ORCDRC_M_sl2_250m", "CRFVOL_M_sl2_250m", "SNDPPT_M_sl2_250m",
                  "PHIHOX_M_sl2_250m", "BDRICM_M_250m"),
         raster = as.list(raster::subset(soilgrids,
                                         c("ORCDRC_M_sl2_250m", "CRFVOL_M_sl2_250m", "SNDPPT_M_sl2_250m",
                  "PHIHOX_M_sl2_250m", "BDRICM_M_250m"))))
  # tibble(group = rep("Terrain", 3),
  #        variable = c("Elevation", "Slope", "Topographic Wetness Index"),
  #        units = c("m", "°", "ln(upslope area / tan(slope))"),
  #        source = rep("SRTM DEM v3", 3),
  #        code = c("srtm1v3_slevant", "srtm1v3_slevant_slope",
  #                         "srtm1v3_slevant_twi"))
)

pred %>%
  dplyr::select(-raster) %>% 
  group_by(group) %>% 
  gt()
```

@@

### Combining predictor variables

These datasets come with different resolutions.
To use them all in the same analysis, we resample them all to the finest dataset (WorldClim, 30 arc-seconds).

```{r predictor-resample}
pred$raster <- lapply(pred$raster, raster::resample, y = bioclim, method = "ngb")
```

## Model construction

@@

```{r predictor-extraction}
# TODO: refactor to use pred
bioclim %>% 
  raster::extract(as_Spatial(occ)) %>% 
  {cbind(occ, .)} %>% 
  st_as_sf(crs = st_crs(occ)) ->
  occ

soilgrids %>% 
  raster::extract(as_Spatial(occ)) %>% 
  {cbind(occ, .)} %>% 
  st_as_sf(crs = st_crs(occ)) ->
  occ
```

@@

```{r response-curves-fit, warning=FALSE, message=FALSE}
# Summarise occurrences on a grid
palestine %>% 
  as_Spatial() %>% 
  raster::rasterize(bioclim) ->
  palestine_raster

occ %>% 
  filter(occurrence == TRUE) %>% 
  group_by(taxon) %>% 
  do(occ_grid = raster_burn(., palestine_raster)) ->
  occ_grids

# Fit poly(1:3) to each taxon/predictor combination
expand.grid(taxon = unique(occ$taxon),
            predictor = pred$code,
            polyn = 1:3) %>% 
  group_by(taxon, predictor, polyn) %>% 
  do(
    # TODO: It works but I hate it.
    fit = model_response(unlist(occ_grids[occ_grids$taxon == .$taxon, "occ_grid"])[[1]],
                         unlist(pred[pred$code == .$predictor, "raster"])[[1]],
                         .$polyn)
  ) %>% 
  glance(fit) -> 
  response_glm
```

```{r response-curves-table, echo=FALSE}
# response_glm %>% 
#   spread(polyn, aic) %>% 
#   rename(aic_poly1 = `1`, aic_poly2 = `2`, aic_poly3 = `3`) %>% 
#   gt(rowname_col = "predictor", groupname_col = "taxon") %>% 
#   tab_stubhead_label("predictor")
```

@@

```{r response-curves-optimal}
response_glm %>%
  group_by(taxon, predictor) %>%
  mutate(optimum = polyn == which.min(AIC)) ->
  response_glm

# TODO
# Re-fit and plot optimal 
# model_response(occ_grids$occ_grid[[1]], pred$raster[[1]], 2) %>% predict(data.frame(pred = -72:109, type = "response")) %>% plot()
# Need to dplyrise this – see broom::augment()
# response_glm %>% 
#   filter(optimum == TRUE) %>% 
#   slice(1:3) %>% 
#   do(
#     fit = model_response(unlist(occ_grids[occ_grids$taxon == .$taxon, "occ_grid"])[[1]],
#                          unlist(pred[pred$code == .$predictor,"raster"])[[1]],
#                          .$polyn)
#   ) 
```

### Model fitting

@@

```{r model_fit, warning=FALSE, message=FALSE}
# TODO: use optimal polynomial order
occ %>% 
  group_by(taxon) %>% 
  remove_missing(vars = c("bio1", "bio2", "bio3", "bio4", "bio5", "bio6", 
                          "bio7", "bio8", "bio9", "bio10", "bio11", "bio12", 
                          "bio13", "bio14", "bio15", "bio16", "bio17", "bio18", 
                          "bio19", "BDRICM_M_250m", "CRFVOL_M_sl2_250m", 
                          "ORCDRC_M_sl2_250m", "PHIHOX_M_sl2_250m", 
                          "SNDPPT_M_sl2_250m")) %>% 
  do(model = glm(occurrence ~ 
                   poly(bio6, 2) +
                   poly(bio12, 2) +
                   poly(bio13, 2) +
                   # poly(bio14, 2) +
                   poly(bio15, 2) +
                   poly(BDRICM_M_250m, 2) +
                   poly(CRFVOL_M_sl2_250m, 2) +
                   poly(ORCDRC_M_sl2_250m, 2) + 
                   poly(PHIHOX_M_sl2_250m, 2) + 
                   poly(SNDPPT_M_sl2_250m, 2),
                 data = .)) ->
  models

```

## Predicting occurrences
### Current conditions

@@

```{r pred_current}
models %>% 
  do(taxon = .$taxon,
     model = .$model,
     pred_current = raster::predict(object = raster::stack(pred$raster),
                                    model = .$model)) ->
  models
```

```{r pred_current_plot, echo=FALSE}
raster::stack(models$pred_current) %>%
  set_names(models$taxon) %>% 
  plot()
```

### Binning

To more easily interpret these predictions, we could convert the raw probability scores into confidence intervals, based on the proportion of known occurrences it successfully predicts.

```{r binning}
raster::stack(models$pred_current) %>%
  set_names(paste0("pred_current_", models$taxon)) %>% 
  raster::extract(as_Spatial(occ)) %>% 
  {cbind(occ, .)} %>% 
  st_as_sf(crs = st_crs(occ)) ->
  occ2
```

>>>>

---
title: Ecological niche models of plant species exploited by Epipalaeolithic foragers
  in the *Badia* (c. 14500 BP)
output:
  pdf_document: 
    latex_engine: xelatex
  html_document:
    df_print: paged
---

```{r packages, include=FALSE}
library("tidyverse")
library("magrittr")
library("sf")
library("stars")
library("parallel")
library("broom")
library("dismo")
library("rnaturalearth")
library("gt")
library("viridis")
library("BadiaPaleoFloraENM")
```

```{r defs, include=FALSE}
latlong <- 4326
utm37n <- 32637
```

@@

```{r regions, include=FALSE}
sw_asia <- extent(c(30, 50, 25, 40))
s_levant <- extent(c(34, 39, 29, 34))

ne_countries(scale = 10, country = c("Israel", "Palestine"), returnclass = "sf") %>% 
  st_union() ->
  palestine

read_sf("data/raw_data/ne/ne_10m_land.shp")[,"featurecla"] %>%
  st_crop(sw_asia) -> 
  land

rivers <- read_sf("data/raw_data/ne/ne_10m_rivers_lake_centerlines_scale_rank.shp") %>% 
  st_crop(s_levant)

water <- rbind(read_sf("data/raw_data/ne/ne_10m_ocean.shp")[,"featurecla"],
               read_sf("data/raw_data/ne/ne_10m_lakes.shp")[,"featurecla"])
```

```{r regions-map, echo=FALSE}
ggplot() + 
  geom_sf(data = palestine, fill = viridis(2)[2], colour = NA) +
  geom_sf(data = rivers, colour = "#999999") +
  geom_sf(data = water, fill = "#eeeeee", colour = "#999999") +
  coord_sf(xlim = c(xmin(s_levant), xmax(s_levant)),
           ylim = c(ymin(s_levant), ymax(s_levant))) +
  theme_bw()
```

## Background

### Plant exploitation in the Epipalaeolithic Southern Levant

### Ecological niche modelling

## Data

### Occurrences

@@

```{r occurrence-data, include=FALSE}
# Download or read data from GBIF
# TODO: T. boeoticum ? (equate with T. urartu?)
gbif_path <- here::here("analysis/data/raw_data/gbif")
occ <- bind_rows(
  fetch_gbif("Hordeum", "spontaneum*", ext = sw_asia, gbif_path),
  fetch_gbif("Bolboschoenus", "glaucus*", ext = sw_asia, gbif_path),    
  fetch_gbif("Bolboschoenus", "maritimus*", ext = sw_asia, gbif_path),    
  fetch_gbif("Scirpus", "maritimus*", ext = sw_asia, gbif_path),
  fetch_gbif("Triticum", "urartu*", ext = sw_asia, gbif_path),
  fetch_gbif("Triticum", "dicoccoides*", ext = sw_asia, gbif_path),
  fetch_gbif("Stipa", "*", ext = sw_asia, gbif_path),
  fetch_gbif("Zilla", "spinosa*", ext = sw_asia, gbif_path),
  fetch_gbif("Vitex", "*", ext = sw_asia, gbif_path),
  fetch_gbif("Fraxinus", "*", ext = sw_asia, gbif_path)
)

# Code taxonomic groups for analysis
occ %>% 
  mutate(taxon = case_when(genus == "Hordeum" ~ "Hordeum spontaneum",
                           genus == "Bolboschoenus" ~ "Bolboschoenus sp.",
                           species == "Triticum urartu" ~ "Triticum urartu",
                           species == "Triticum dicoccoides" ~ "Triticum dicoccoides",
                           genus == "Stipa" ~ "Stipa sp.",
                           genus == "Zilla" ~ "Zilla spinosa",
                           genus == "Vitex" ~ "Vitex sp.",
                           genus == "Fraxinus" ~ "Fraxinus sp.")) %T>%
  {if(any(is.na(.$taxon))) stop("Taxon contains missing values!")} ->
  occ

@@

It's interesting to map the current distribution of these plants.
```{r occurrence-map, echo=FALSE}
ggplot(occ, aes(x = lon, y = lat)) +
  geom_sf(data = water, inherit.aes = FALSE, fill = "#eeeeee", colour = "#999999") +
  geom_sf(data = occ, shape = 1, size=0.5) +
  coord_sf(xlim = c(xmin(sw_asia), xmax(sw_asia)),
           ylim = c(ymin(sw_asia), ymax(sw_asia))) +
  facet_wrap(~taxon) +
  theme_bw() +
  theme(axis.text = element_blank(),
        axis.ticks = element_blank(),
        axis.title = element_blank())
```

### Pseudo-absences

Occurrence data only tells us where a species is present;
there is no definitive information on where the species is *not* found.
We therefore need to generate random background points or "pseudo-absences" to feed to the model.
There are several ways to do this. 
We follow the advice of Barbet-Massin et al. (2012) for regression-based species distribution models and use a large (~10000) random sample of points, weighted equally against the presences in the regression.

```{r psabsences, include=FALSE}
n_psab <- 10000

occ <- as_tibble(occ)
occ$occurrence <- TRUE
for(taxon in unique(occ$taxon)) {
  occ <- bind_rows(occ, generate_psabsences(palestine, taxon = taxon,
                                            occurrence = FALSE, N = n_psab))
}
occ <- sf::st_as_sf(occ, crs = 4326)
rm(taxon)

# Weights
occ %>% 
  group_by(taxon, occurrence) %>% 
  mutate(weight = 1 + (n_psab / n() - 1)) ->
  occ
```

### Environmental data

#### Current climate

@@

```{r bioclim-table, echo=FALSE}
tibble(biovar = paste0("bio", 1:19),
       definition = c("Mean annual temperature",
                      "Mean diurnal range",
                      "Isothermality",
                      "Temperature seasonality",
                      "Maximum temperature of warmest month",
                      "Minimum temperature of coldest month",
                      "Annual temperature range",
                      "Mean temperature of wettest quarter",
                      "Mean temperature of driest quarter",
                      "Mean temperature of warmest quarter",
                      "Mean temperature of coldest quarter",
                      "Total annual precipitation",
                      "Precipitation of wettest month",
                      "Precipitation of driest month",
                      "Precipitation seasonality",
                      "Precipitation of wettest quarter",
                      "Precipitation of driest quarter",
                      "Precipitation of warmest quarter",
                      "Precipitation of coldest quarter"),
       formula = c("",
                   "",
                   "(bio2 / bio7) * 100",
                   "sd(temp) * 100",
                   "",
                   "",
                   "bio5 - bio6",
                   "",
                   "",
                   "",
                   "",
                   "",
                   "",
                   "",
                   "",
                   "",
                   "",
                   "",
                   "")
) %>%
  gt()
```

Here the current bioclimatic data is from CHELSA (<http://chelsa-climate.org/bioclim/>).

```{r bioclim-data, include=FALSE}
fetch_paleoclim("cur", "2_5m", s_levant,
                "./data/derived_data/paleoclim/paleoclim_cur.grd") %>% 
  st_as_stars() ->
  bioclim
```

#### Palaeoclimate

```{r paleoclim-data, include=FALSE}
tibble(
  period_code = c("eh", "yds", "ba", "hs1", "lgm"),
  period = c("Early Holocene", "Younger Dryas", "Bølling-Allerød", "Heinrich 1", "Last Glacial Maximum"),
  raster = c(
    fetch_paleoclim("eh", "2_5m", s_levant, 
                    "./data/derived_data/paleoclim/paleoclim_eh.grd"),
    fetch_paleoclim("yds", "2_5m", s_levant, 
                    "./data/derived_data/paleoclim/paleoclim_yds.grd"),
    fetch_paleoclim("ba", "2_5m", s_levant, 
                    "./data/derived_data/paleoclim/paleoclim_ba.grd"),
    fetch_paleoclim("hs1", "2_5m", s_levant, 
                    "./data/derived_data/paleoclim/paleoclim_hs1.grd"),
    fetch_paleoclim("lgm", "2_5m", s_levant, 
                    "./data/derived_data/paleoclim/paleoclim_lgm.grd")
  )
) %>% 
  mutate(raster = st_as_stars(raster)) ->
  paleoclim
```

### Soil

@@

```{r soil-data, include=FALSE}
soilgrids <- raster::stack(list.files("data/raw_data/soilgrids", ".*tif",
                                      full.names = TRUE))
```

@@

There are also relevent aspects of the terrain that can be easily derived from a digital elevation model.
We'll use NASA-JPL's SRTM v3 3 arc-second resolution DEM, smoothed with Sun's denoising algorithm [TODO: CITE].
Slope was calculated from this surface using GDAL 3.0.0 and the topographic wetness index (TWI) using the r.topidx module of GRASS 7.8.
The DEM was reprojected to a conformal projection (Jordan Transverse Mercator, EPSG:19995) for these calculations, using a bilinear interpolation, then back to geographic coordinates.

```{r terrain-data, include=FALSE}
raster::stack(paste0("./data/raw_data/srtm/",
                     c("srtm3v3s_slevant.tif",
                       "srtm3v3s_slevant_slope.tif",
                       "srtm3v3s_slevant_twi.tif"))) %>% 
  set_names(c("elev", "slope", "twi")) ->
  terrain
```

## Model

### Selection of predictors

@@

```{r review-table, echo=FALSE}
read_csv("data/raw_data/flora_review.csv") %>% 
  group_by(Taxon) %>% 
  # TODO: add refs
  dplyr::select(-Reference) %>% 
  gt()
```

@@

```{r predictor-table, echo=FALSE}
pred <- bind_rows(
  tibble(group = rep("Climate", 4),
         time = rep("now", 4),
         variable = c("Minimum temperature of coldest month",
                      "Total annual precipitation", "Precipitation of wettest month",
                      "Precipitation seasonality"),
         units = c("°C ×10", "mm", "mm", "σ/μ"),
         source = rep("PaleoClim 1.0", 4),
         code = c("bio_6", "bio_12", "bio_13", "bio_15"),
         raster = as.list(raster::subset(bioclim,
                                         c("bio_6", "bio_12", "bio_13",
                                           "bio_15")))),
  tibble(group = rep("Soil", 5),
         time = rep(NA, 5),
         variable = c("Organic carbon content", "Coarse fragment content",
                      "Sand content", "Acidity", "Depth to bedrock"),
         units = c("g per kg fine earth", "% volume", "% mass", "pH ×10 in H2O",
                   "cm (up to 200 cm)"),
         source = rep("SoilGrids250m", 5),
         code = c("ORCDRC_M_sl2_250m", "CRFVOL_M_sl2_250m", "SNDPPT_M_sl2_250m",
                  "PHIHOX_M_sl2_250m", "BDRICM_M_250m"),
         raster = as.list(raster::subset(soilgrids,
                                         c("ORCDRC_M_sl2_250m", "CRFVOL_M_sl2_250m", "SNDPPT_M_sl2_250m",
                  "PHIHOX_M_sl2_250m", "BDRICM_M_250m")))),
  tibble(group = rep("Terrain", 3),
         time = rep(NA, 3),
         variable = c("Elevation", "Slope", "Topographic Wetness Index"),
         units = c("m", "°", "ln(upslope area / tan(slope))"),
         source = rep("SRTM DEM v3", 3),
         code = c("srtm1v3s_slevant", "srtm1v3s_slevant_slope",
                  "srtm1v3s_slevant_twi"),
         raster = as.list(terrain))
)

# PaleoClim
paleoclim %>% 
  transmute(group = "Climate", time = period_code, raster = raster) %>% 
  mutate(raster = map(raster, ~raster::unstack(.))) %>% 
  unnest() %>% 
  mutate(code = unlist(map(raster, names))) %>% 
  filter(code %in% c("bio_6", "bio_12", "bio_13", "bio_15")) %>% 
  {bind_rows(pred, .)} ->
  pred

pred %>%
  filter(time == "now" | is.na(time)) %>% 
  dplyr::select(-raster, -time) %>% 
  group_by(group) %>% 
  gt() %>% 
  cols_label(variable = "Variable", units = "Units", source = "Dataset",
             code = "Source code")
```

@@

These datasets come with different resolutions.
To use them all in the same analysis, we resample them all to the finest dataset (the SRTM terrain variables, at 3 arc-second resolution).

```{r predictor-resample, include=FALSE}
# Parallelisation is Linux and Mac only.
# 
pred$raster <- mclapply(pred$raster, raster::resample, y = terrain, method = "ngb",
                        mc.cores = detectCores())
# pred$raster <- lapply(pred$raster, raster::resample, y = terrain, method = "ngb")
```

@@

```{r predictor-extraction, include=FALSE}
pred %>% 
  filter(time == "now" | is.na(time)) %>% 
  magrittr::extract2("raster") %>% 
  stack() %>% 
  raster::extract(as_Spatial(occ)) %>% 
  {cbind(occ, .)} %>% 
  st_as_sf(crs = st_crs(occ)) ->
  occ
```

@@

```{r response-curves-fit, include=FALSE}
occ %>% 
  gather("pred_var", "pred_val", bio_6:twi) %>% 
  # NAs from gaps in the predictor data.
  drop_na(pred_val) %>%
  group_by(taxon, pred_var) %>% 
  nest() %>% 
  expand(taxon, pred_var, data, polyn = 1:3) %>%
  mutate(
    response_model = paste0("occurrence ~ poly(pred_val, ", polyn, ")"),
    response_fit = map2(data, response_model,
                        ~ glm(as.formula(.y),
                              family = "binomial",
                              data = .x,
                              weights = weight)),
    response_glance = map(response_fit, glance)
  ) ->
  response_models

# Select model with lowest AIC for each taxon–predictor combination
response_models %>%
  unnest(response_glance) %>% 
  group_by(taxon, pred_var) %>% 
  filter(AIC == min(AIC)) ->
  response_models
```

```{r response-curves-table, echo=FALSE}
# response_models %>% 
#   spread(polyn, aic) %>% 
#   rename(aic_poly1 = `1`, aic_poly2 = `2`, aic_poly3 = `3`) %>% 
#   gt(rowname_col = "predictor", groupname_col = "taxon") %>% 
#   tab_stubhead_label("predictor")
```

@@

### Logistic regression

@@

```{r model_fit, include=FALSE}
occ %>% 
  drop_na(bio_6:twi) %>% 
  group_by(taxon) %>% 
  nest() %>% 
  mutate(
    model = map_chr(taxon, ~ paste0(
      "occurrence ~ ",
      str_c(paste0("poly(",
                   filter(response_models, taxon == .x)$pred_var,
                   ", ",
                   filter(response_models, taxon == .x)$polyn,
                   ")"),
            collapse = " + "))),
    fit = map2(model, data,
               ~ glm(as.formula(.x),
                     family = "binomial",
                     data = .y,
                     weights = weight)),
    glance = map(fit, glance)
  ) ->
  enm
```

## Results
### Current conditions

@@

```{r predictions, include=FALSE}
# Slow
profvis::profvis({
enm %>% 
  mutate(pred_now = map(fit, 
                        ~raster::predict(
                          raster::stack(
                            filter(pred, time == "now" | is.na(time))$raster),
                          .))) %>% 
  mutate(pred_eh = map(fit, 
                        ~raster::predict(
                          raster::stack(
                            filter(pred, time == "eh" | is.na(time))$raster),
                          .))) %>% 
  mutate(pred_yds = map(fit, 
                         ~raster::predict(
                           raster::stack(
                             filter(pred, time == "yds" | is.na(time))$raster),
                           .))) %>% 
  mutate(pred_ba = map(fit, 
                        ~raster::predict(
                          raster::stack(
                            filter(pred, time == "ba" | is.na(time))$raster),
                          .))) %>% 
  mutate(pred_hs1 = map(fit, 
                         ~raster::predict(
                           raster::stack(
                             filter(pred, time == "hs1" | is.na(time))$raster),
                           .))) %>% 
  mutate(pred_lgm = map(fit, 
                         ~raster::predict(
                           raster::stack(
                             filter(pred, time == "lgm" | is.na(time))$raster),
                           .))) %>% 
  mutate(pred_now = map(pred_now, logit_to_p),
         pred_eh = map(pred_eh, logit_to_p),
         pred_yds = map(pred_yds, logit_to_p),
         pred_ba = map(pred_ba, logit_to_p),
         pred_hs1 = map(pred_hs1, logit_to_p),
         pred_lgm = map(pred_yds, logit_to_p)) ->
  enm
})
```

```{r pred_current_plot, echo=FALSE}
raster::stack(enm$pred_now) %>%
  set_names(enm$taxon) %>% 
  plot()
```

```{r pred_yds_}

## Discussion
